---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gemmaAPI

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/gemmaAPI)](https://CRAN.R-project.org/package=gemmaAPI)
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/gemmaAPI.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/gemmaAPI)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://www.tidyverse.org/lifecycle/#stable)
<!-- badges: end -->

This is an R wrapper for [Gemma](http://gemma.msl.ubc.ca)’s restful [API](https://gemma.msl.ubc.ca/resources/restapidocs/).

To cite Gemma, please use: [Lim, N. et al., Curation of over 10 000 transcriptomic studies to enable data reuse, Database, 2021.](https://doi.org/10.1093/database/baab006)

## Installation

You can install the released version of gemmaAPI from [CRAN](https://CRAN.R-project.org) with:

```{r, eval = FALSE}
install.packages("gemmaAPI")
```

And the development version from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("gaborcsardi/async")
devtools::install_github("jsicherman/GemmAPI")
```

## Usage

For help on using the Gemma API wrapper, please see [the REST API docs](https://gemma.msl.ubc.ca/resources/restapidocs/) or the [vignette](vignettes/Usage.md).
