---
title: "A meta analysis on effects of Parkinson's Disease using Gemma.R"
author:
- name: B. Ogan Mancarci
  affiliation: Michael Smith Laboratories, University of British Columbia, Vancouver, Canada
package: gemma.R
output:
    BiocStyle::html_document
header-includes:
vignette: >
    %\VignetteIndexEntry{A meta analysis on effects of Parkinson's Disease using Gemma.R}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
# Prevent certificate issues for GitHub actions
options(gemma.SSL = FALSE,gemma.memoised = TRUE)
# options(gemma.API = "https://dev.gemma.msl.ubc.ca/rest/v2/")
knitr::opts_chunk$set(
    comment = ""
)
```

```{r setup, message = FALSE}
library(gemma.R)
library(dplyr)
library(poolr)
library(magrittr)
library(ggplot2)
```

Gemma.R contains a large number of datasets representing a wide array of
conditions. It is therefore a useful tool for running meta-analysis
based on publicly available datasets.

In this example we will search Gemma for datasets comparing healthy
controls and patients with Parkinson's disease, and run a meta-analysis
to discover genes consistently found to be related to the disease.

# Querying datasets of interest

`search_datasets` function can be used to find datasets of interest,
either using ontology terms like
"<http://purl.obolibrary.org/obo/DOID_14330>" or plain text.

```{r}
# getting all resulting datasets using limit and offset arguments
result <- search_datasets('http://purl.obolibrary.org/obo/DOID_14330',limit = 1)
result_count <- attributes(result)$totalElements

results <- lapply(seq(0,result_count,100),function(i){
    search_datasets('http://purl.obolibrary.org/obo/DOID_14330',
                    limit = 100,offset = i)
}) %>% do.call(rbind,.)

results %>% select(experiment.ShortName,experiment.Name,taxon.Name) %>% head
```

A similar search is also possible using plain text

```{r}
result <- search_datasets('parkinsons',limit = 1)
result_count <- attributes(result)$totalElements


results_plain <- lapply(seq(0,result_count,100),function(i){
    search_datasets('parkinsons',limit = 100,offset = i)
}) %>% do.call(rbind,.)


results_plain %>% 
    select(experiment.ShortName,experiment.Name,taxon.Name) %>% head
```

These two queries return similar but not-identical results.

```{r}
# plain text returns more results than the annotation term
results_plain %>% nrow

results %>% nrow

# annotation term query is a subset of plain text query
length(intersect(
    results_plain$experiment.ShortName,
    results$experiment.ShortName))
```

We will be using the results from the annotation term search in the rest
of this analysis since plain text search will match with any mention of
the disease if it is included in the study description.

# Filtering the datasets for suitability

While we know that all the resulting datasets were annotated for the
term for Parkinson's Disease, we currently do not know how many of them
are comparisons of healthy controls and patients with Parkinson's
disease. Looking at the sample above, quite a few of them aren't from
humans in the first place. To find the datasets that are suitable for
our purposes we need to first remove all non-human datasets, then ensure
comparison performed is on Parkinson's disease patients and controls

```{r}
# isolate human datasets
human_results <- results %>% filter(taxon.Name == 'human')

nrow(human_results)
```

Next step is to ensure the datasets we are selecting are not affected by
batch confounds. Gemma internally handles batch correction if batch
information is available for the dataset. We will be looking at
`geeq.batchEffect` column. As explained in the `search_datasets`
documentation, this column will be set to -1 for datasets where batch
confound is detected, 0 for datasets without available batch information
and to 1 if the data is clear of batch confounds.

```{r}
human_results <- human_results %>% filter(geeq.batchEffect == 1)
```

We now want to ensure that the differential expressions we analyze
compare control and Parkinson's Disease patients. This information is
available via `get_dataset_differential_expression_analyses` which
returns the experimental groups for differential expression analyses
performed for the dataset. The columns we are primarily interested in
are `baseline.factorValue`/`baseline.factorValueURI` which typically
records the control group of the differential expression analysis and
`experimental.factorValue`/`experimental.factorValueURI` which typically
records the test case

```{r}
experiment_contrasts <- human_results$experiment.ShortName %>% 
    lapply(function(x){
        out = get_dataset_differential_expression_analyses(x)
        }) %>% do.call(rbind,.)

experiment_contrasts %>%
    select(experiment.ID,result.ID,
           baseline.factorValue,
           baseline.factorValueURI,
           experimental.factorValue,
           experimental.factorValueURI) %>% head
```

The factor we are interested in is Parkinson's Disease in
`experimental.factorValue` or
"<http://purl.obolibrary.org/obo/DOID_14330>" in
`experimental.factorValueURI`.

```{r}
parkin_contrasts <- experiment_contrasts %>% 
    filter(experimental.factorValueURI == "http://purl.obolibrary.org/obo/DOID_14330")

parkin_contrasts %>%
    select(experiment.ID,
           result.ID,
           baseline.factorValue,
           baseline.factorValueURI,
           experimental.factorValue,
           experimental.factorValueURI) %>% 
    head()
```

Now that we have our relevant contrasts, we can download them using
`get_differential_expression_values`. This function can be used download
differential expression fold change and p values, either using the
experiment name/ids or more specifically using the result.IDs

```{r}
differentials <- parkin_contrasts$result.ID %>% lapply(function(x){
    # take the first and only element of the output. the function returns a list 
    # because single experiments may have multiple resultSets. Here we use the 
    # resultSet argument to directly access the results we need
    get_differential_expression_values(resultSet = x)[[1]]
})



# some datasets might not have all the advertised differential expression results
# calculated due to a variety of factors. here we remove the empty differentials

missing_contrasts <- differentials %>% sapply(nrow) %>% {.==0}
differentials <- differentials[!missing_contrasts]
parkin_contrasts <- parkin_contrasts[!missing_contrasts,]

```

# Getting the p-values for the condition comparison

`differentials` is now a list of data frames containing the differential
expression information. To run a meta-analysis, we need to p values for
the genes from the relevant contrasts.

```{r}

condition_diffs <- seq_along(differentials) %>% lapply(function(i){
    # iterate over the differentials
    diff = differentials[[i]]
    # get the contrast information about the differential
    contrast = parkin_contrasts[i,]
    p_vals = diff[[paste0('contrast_',contrast$contrast.id,"_pvalue")]]
    log2fc = diff[[paste0('contrast_',contrast$contrast.id,"_log2fc")]]
    genes = diff$GeneSymbol
    
    
    data.frame(genes,p_vals,log2fc)
})

# we can use result.IDs and contrast.IDs to uniquely name this. 
# we add the experiment.id for readability
names(condition_diffs) = paste0(parkin_contrasts$experiment.ID,'.',
                                parkin_contrasts$result.ID,'.',
                                parkin_contrasts$contrast.id)

condition_diffs[[1]] %>% head
```

# Combining the acquired p values

We will now apply Fisher's method to combine p values to find our genes
of interest

First step is to identify which genes are available in our results

```{r}
all_genes <- condition_diffs %>% lapply(function(x){
    x$genes %>% unique
}) %>% unlist %>% table

# we will remove any gene that doesn't appear in at least half the results
all_genes <- all_genes[all_genes>(length(condition_diffs)/2)]
all_genes <- names(all_genes)

# remove any probesets matching multiple genes. gemma separates these by using "|"
all_genes <- all_genes[!grepl("|",all_genes,fixed = TRUE)]

# remove the "". This comes from probesets not aligned to any genes
all_genes <- all_genes[all_genes != ""]
all_genes %>% head
```

Now we can run the test on every gene, followed by a multiple testing correction.

```{r, fisher}
fisher_results <- all_genes %>% lapply(function(x){
    p_vals <- condition_diffs %>% sapply(function(y){
        # we will resolve multiple probesets by taking the minimum p value for
        # this example
        out = y[y$genes == x,]$p_vals
        if(length(out) == 0 ||all(is.na(out))){
            return(NA)
        } else{
            return(min(out))
        }
    })
    
    fold_changes <- condition_diffs %>% sapply(function(y){
        pv = y[y$genes == x,]$p_vals
        if(length(pv) == 0 ||all(is.na(pv))){
            return(NA)
        } else{
            return(y[y$genes == x,]$log2fc[which.min(pv)])
        }
    })
    
    median_fc = fold_changes %>% na.omit() %>% median
    names(median_fc) = 'Median FC'
    
    combined = p_vals %>% na.omit() %>% fisher() %$% p
    names(combined) = 'Combined'
    c(p_vals,combined,median_fc)
}) %>% do.call(rbind,.)
fisher_results <- as.data.frame(fisher_results)
rownames(fisher_results) = all_genes

fisher_results[,'Adjusted'] <- p.adjust(fisher_results[,'Combined'],
                                        method = 'fdr')

fisher_results %>%
    arrange(Adjusted) %>% 
    select(Combined,Adjusted,`Median FC`) %>% 
    head
```
We end up with quite a few differentially expressed genes

```{r}
sum(fisher_results$Adjusted<0.05) # FDR<0.05
nrow(fisher_results) # number of all genes
```

A quick sanity check would be to look at markers of dopaminergic cell types and how they rank compared to other genes.
Parkinson's disease is a neurodegenerative disorder, leading to death of dopaminergic cells. We should expect them to show up in our
results.

```{r}
# markers are taken from https://www.eneuro.org/content/4/6/ENEURO.0212-17.2017
dopa_markers <-  c("ADCYAP1", "ATP2B2", "CACNA2D2", 
"CADPS2", "CALB2", "CD200", "CDK5R2", "CELF4", "CHGA", "CHGB", 
"CHRNA6", "CLSTN2", "CNTNAP2", "CPLX1", "CYB561", "DLK1", "DPP6", 
"ELAVL2", "ENO2", "GABRG2", "GRB10", "GRIA3", "KCNAB2", "KLHL1", 
"LIN7B", "MAPK8IP2", "NAPB", "NR4A2", "NRIP3", "HMP19", "NTNG1", 
"PCBP3", "PCSK1", "PRKCG", "RESP18", "RET", "RGS8", "RNF157", 
"SCG2", "SCN1A", "SLC12A5", "SLC4A10", "SLC6A17", "SLC6A3", "SMS", 
"SNCG", "SPINT2", "SPOCK1", "SYP", "SYT4", "TACR3", "TENM1", 
"TH", "USP29")

fisher_results %>% 
    arrange(Combined) %>% 
    rownames %>%
    {.%in% dopa_markers} %>%
    which %>% 
    hist(main = 'Rank distribution of dopaminergic markers')
```

We can see that the dopaminergic markers tend to have high ranks in our results.
