---
title: "Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gemmaAPI)
library(dplyr)
```

# Basic usage

The Gemma API wrapper is very straightforward to use and contains a function for each entry in the REST API docs, [here](https://gemma.msl.ubc.ca/resources/restapidocs/).

## Downloading basic information for a dataset

The first endpoint example for the REST API doc corresponds to a simple case: getting basic information about a single dataset in Gemma, such as [GSE2018](https://gemma.msl.ubc.ca/expressionExperiment/showExpressionExperiment.html?id=1). We can do this in one of two ways using the API:

1.  Using the dataset's shortname (GSE2018), or
2.  Using the dataset's identifier (1, as you can see in the URL)

Each case is functionally identical in the API wrapper.

```{r}
glimpse(getDatasets(dataset = 1))


glimpse(getDatasets(dataset = 'GSE2018'))
```

Most endpoints also allow you to query multiple objects at the same time, for example:

```{r}
glimpse(getDatasets(dataset = c('GSE2018', 'GSE25136')))
```

# Asynchronous requests

# Memoised results

## Clearing cache

# File outputs

# Raw data
